Preventing a Nuclear ISIL: \\Recommendations for United States Policy

### 2015-05-11 draft.tex
Finalized list of acronyms.

AAPA - American Association of Port Authorities

@@

WMD - Weapons of Mass Destruction

@@

Following the interception of SNM or weapon components, careful analysis of the composition, material structure, and relative sophistication of the physics package will allow forensic investigators to estimate the age of the device, the techniques which were utilized in order to create the physics package, and the most probable country of origin \cite{Glaser2008}. In the U.S., a number of organizations would be involved in these assessments. The National Technical Nuclear Forensics Center (NTNFC) is responsible for ensuring the nation's nuclear forensics capability.  The NTNFC is supported by the Department of Homeland Security (DHS), Department of Justice (DoJ), Federal Bureau of Investigation (FBI), Department of Defense (DoD), Department of Energy (DoE), Department of State (DoS), and the Office of the Director of National Intelligence (ODNI) \cite{DepartmentOfHomelandSecurit}.

@@

Two of the relevant governmental organizations regulating domestic borders and security are the Federal Bureau of Investigation (FBI) and the Department of Homeland Security (DHS). Part of the responsibility of these organizations is ensuring the security of the U.S. from both domestic and foreign threats. They coordinate intelligence gathering, threat-response, and evidence collection in support of the justice system. With these goals in mind, a significant portion of their work deals with the monitoring of borders. The Department of Homeland Security alone commands more than 180,000 personnel, operating such sub-groups as the U.S. Coast Guard, U.S. Customs and Border Protection (CBP), and the Immigration and Customs Enforcement (ICE) \cite{Robb2005}. One of the most relevant DHS sub-groups is the Domestic Nuclear Detection Office (DNDO). The DNDO is responsible for the development and deployment of radiation and detection equipment to assist any other federal agency, including the other DHS sub-organizations and the FBI \cite{UnitedStatesGovernmentAccountabilityOffice2013}.

@@

The CPB/AAPA RPM plan outlines a phased implementation approach targeting locations around the U.S. based on their assessed risk. The plan additionally details collaboration with the Pacific Northwest National Lab (PNNL), which is responsible for the development of cutting-edge, harmless, and passive detection systems (illustrated in \autoref{fig:RPM_designs}). In 2005, RPM systems had already been deployed at 14 terminals, with another 18 in construction, and 27 in final design \cite{Simmons2005}.

@@

In fact, diplomatic and technical solutions are the bulk of the policy proposals put forth in this report.  In theory, this is a great approach as it would not be difficult to influence the American people or U.S. government to support efforts to combat terrorism, nuclear or non-nuclear, in the post-9/11 environment. Since 9/11, the U.S. has largely adopted a \enquote{1\% Doctrine} where all threats with even a 1\% likelihood should be eliminated \cite{pollack2014unthinkable}. This, coupled with a general reluctance of the American populace to question the U.S. military or anything posed as a national security issue, has given de facto carte blanche to all \enquote{good idea fairies} to promote their particular product or idea in the name of improving national security \cite{Fallows2015}. A relevant example to the proposed policies in this report is the history and effectiveness of radiation portal monitors (RPMs) and the Advance Spectroscopic Portals (ASPs).

@@

The most substantial effect of an RDD is the spread of fear, panic, economic disruption, and the cost associated with cleaning up a contaminated area. These devices are unlikely to result in as many deaths as a fission or thermonuclear device due to the short duration of radiation exposure and the diffuse distribution of the dispersed materials. \autoref{tab:health_effects} from the U.S. Environmental Protection Agency shows the level of exposure required to produce various health symptoms and the time from acute exposure to symptom onset \cite{USEPA1999}. Department of Homeland Security (DHS) and local law enforcement crisis management plans to evacuate crowds following an RDD detonation and the subsequent radiation cleanup will drastically reduce the number of people who receive radiation doses large enough to cause short term health effects.

@@

Intelligence agencies gather information from physical and technical observations, conversations, searches, and human operations. These sources are grouped into Signals Intelligence (SIGINT), Imagery Intelligence (IMINT), Measurement and Signature Intelligence (MASINT), Human-Source Intelligence (HUMINT), Open-Source Intelligence (OSINT) and Geospatial Intelligence (GEOINT) \cite{Intelligen}. The National Security Agency (NSA) is responsible for SIGINT, and the NSA collects SIGINT using signal communications between people, machines, and combinations of both. National Geospatial-Intelligence Agency (NGA) is responsible for the collection of IMINT using visual collection methods including cameras, radars, lasers, etc. MASINT is technically and scientifically based intelligence on unique attributes of the desired target. MASINT is the part of the Defense Intelligence Agency (DIA) and can be used to recognize, locate, and describe the target through nuclear, seismic, acoustic, material, radio frequency, and optical sciences. The Department of State (DoS), CIA, Department of Defense (DoD), Federal Bureau of Investigation (FBI), and other agencies use HUMINT sources, more commonly known as spies, to gather data via photography, document transfer, information download, etc. in enemy territory. OSINT relies on publicly available information, and a large synthesis of OSINT is performed at Foreign Broadcast Information Service (FBIS) and the National Air and Space Intelligence Center (NASIC). GEOINT belongs to the NGA and is generally collected from satellites and reconnaissance aircraft.

@@

The UN is the largest international organization that can take actions on the issues of security at a global level. Several divisions in the UN address nuclear security issues including the General Assembly (GA), the Security Council (SC), and the Department of Disarmament Affairs (DDA). The GA can make recommendations to Member States and the SC about situations that are likely to endanger international peace and security. The SC consists of five permanent member states (China, France, Russia, United Kingdom, and the United States) and 10 non-permanent members. The Security Council is responsible for formulating plans for the regulation of armaments. The DDA provides substantive and organizational support to UN Member States for disarmament norm-setting through the work of the GA and other bodies. The DDA's Weapons of Mass Destruction (WMD) Branch supports multilateral efforts to strengthen the international norm on disarmament and non-proliferation of WMD and cooperates with other UN agencies, such as the IAEA.

>>>>

{\parskip=0.5ex AAPA - American Association of Port Authorities

@@

WMD - Weapons of Mass Destruction}

@@

Following the interception of SNM or weapon components, careful analysis of the composition, material structure, and relative sophistication of the physics package will allow forensic investigators to estimate the age of the device, the techniques which were utilized in order to create the physics package, and the most probable country of origin \cite{Glaser2008}. In the U.S., a number of organizations would be involved in these assessments. The National Technical Nuclear Forensics Center (NTNFC) is responsible for ensuring the nation's nuclear forensics capability.  The NTNFC is supported by the Department of Homeland Security (DHS), Department of Justice (DoJ), Federal Bureau of Investigation (FBI), Department of Defense (DoD), Department of Energy (DoE), Department of State (DoS), and the Office of the Director of National Intelligence  \cite{DepartmentOfHomelandSecurit}.

@@

Two of the relevant governmental organizations regulating domestic borders and security are the Federal Bureau of Investigation  and the Department of Homeland Security. Part of the responsibility of these organizations is ensuring the security of the U.S. from both domestic and foreign threats. They coordinate intelligence gathering, threat-response, and evidence collection in support of the justice system. With these goals in mind, a significant portion of their work deals with the monitoring of borders. The Department of Homeland Security alone commands more than 180,000 personnel, operating such sub-groups as the U.S. Coast Guard, U.S. Customs and Border Protection (CBP), and the Immigration and Customs Enforcement (ICE) \cite{Robb2005}. One of the most relevant DHS sub-groups is the Domestic Nuclear Detection Office. The DNDO is responsible for the development and deployment of radiation and detection equipment to assist any other federal agency, including the other DHS sub-organizations and the FBI \cite{UnitedStatesGovernmentAccountabilityOffice2013}.

@@

The CPB/AAPA RPM plan outlines a phased implementation approach targeting locations around the U.S. based on their assessed risk. The plan additionally details collaboration with the Pacific Northwest National Lab, which is responsible for the development of cutting-edge, harmless, and passive detection systems (illustrated in \autoref{fig:RPM_designs}). In 2005, RPM systems had already been deployed at 14 terminals, with another 18 in construction, and 27 in final design \cite{Simmons2005}.

@@

In fact, diplomatic and technical solutions are the bulk of the policy proposals put forth in this report.  In theory, this is a great approach as it would not be difficult to influence the American people or U.S. government to support efforts to combat terrorism, nuclear or non-nuclear, in the post-9/11 environment. Since 9/11, the U.S. has largely adopted a \enquote{1\% Doctrine} where all threats with even a 1\% likelihood should be eliminated \cite{pollack2014unthinkable}. This, coupled with a general reluctance of the American populace to question the U.S. military or anything posed as a national security issue, has given de facto carte blanche to all \enquote{good idea fairies} to promote their particular product or idea in the name of improving national security \cite{Fallows2015}. A relevant example to the proposed policies in this report is the history and effectiveness of radiation portal monitors and the Advance Spectroscopic Portals.

@@

The most substantial effect of an RDD is the spread of fear, panic, economic disruption, and the cost associated with cleaning up a contaminated area. These devices are unlikely to result in as many deaths as a fission or thermonuclear device due to the short duration of radiation exposure and the diffuse distribution of the dispersed materials. \autoref{tab:health_effects} from the U.S. Environmental Protection Agency shows the level of exposure required to produce various health symptoms and the time from acute exposure to symptom onset \cite{USEPA1999}. Department of Homeland Security  and local law enforcement crisis management plans to evacuate crowds following an RDD detonation and the subsequent radiation cleanup will drastically reduce the number of people who receive radiation doses large enough to cause short term health effects.

@@

Intelligence agencies gather information from physical and technical observations, conversations, searches, and human operations. These sources are grouped into Signals Intelligence (SIGINT), Imagery Intelligence (IMINT), Measurement and Signature Intelligence (MASINT), Human-Source Intelligence (HUMINT), Open-Source Intelligence (OSINT) and Geospatial Intelligence (GEOINT) \cite{Intelligen}. The National Security Agency (NSA) is responsible for SIGINT, and the NSA collects SIGINT using signal communications between people, machines, and combinations of both. National Geospatial-Intelligence Agency (NGA) is responsible for the collection of IMINT using visual collection methods including cameras, radars, lasers, etc. MASINT is technically and scientifically based intelligence on unique attributes of the desired target. MASINT is the part of the Defense Intelligence Agency (DIA) and can be used to recognize, locate, and describe the target through nuclear, seismic, acoustic, material, radio frequency, and optical sciences. The Department of State, CIA, Department of Defense, Federal Bureau of Investigation, and other agencies use HUMINT sources, more commonly known as spies, to gather data via photography, document transfer, information download, etc. in enemy territory. OSINT relies on publicly available information, and a large synthesis of OSINT is performed at Foreign Broadcast Information Service (FBIS) and the National Air and Space Intelligence Center (NASIC). GEOINT belongs to the NGA and is generally collected from satellites and reconnaissance aircraft.

@@

The UN is the largest international organization that can take actions on the issues of security at a global level. Several divisions in the UN address nuclear security issues including the General Assembly (GA), the Security Council (SC), and the Department of Disarmament Affairs (DDA). The GA can make recommendations to Member States and the SC about situations that are likely to endanger international peace and security. The SC consists of five permanent member states (China, France, Russia, United Kingdom, and the United States) and 10 non-permanent members. The Security Council is responsible for formulating plans for the regulation of armaments. The DDA provides substantive and organizational support to UN Member States for disarmament norm-setting through the work of the GA and other bodies. The DDA's WMD Branch supports multilateral efforts to strengthen the international norm on disarmament and non-proliferation of WMD and cooperates with other UN agencies, such as the IAEA.
